---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ReliableTrendIndex

<!-- badges: start -->
<!-- badges: end -->

An R package for analysis of the reliability of changes.

## What is a reliable change?  

The idea of a reliable change is that some sequences of scores, which appear to show change (*observed* changes), do not reflect *true* changes. Usually, these sequences of scores are just a sequence of two scores, simplified to a difference score. The point of the ReliableTrendIndex package is that in principle, these sequences can be any length. Regardless of the length of the sequence of observations, only *true* changes matter, in this theory, and anything that is observed but not true change should be treated as if it is noise.  

If you want to identify reliable changes, you are interested in separating those sequences of scores that contain true change from sequences that do not contain true changes. You cannot do this with perfect accuracy. You can only estimate likelihoods of underlying true change states: most commonly, the chance that the observations could occur if there was no true change at all.  

The reliable change index (RCI) separates difference scores based on the chance that a difference at least as large as the observed one could be due to chance alone. It uses an estimate of measurement error from external data to evaluate this. Since the RCI only pertains to a difference score, it only accommodates two observation sequences. This is a limitation if you have more than one observation, since it will ignore the additional observations. 

The reliable trend index (RTI) looks for reliability of trends, rather than change scores. That is, it takes all data points and compares the line of best fit to a perfectly unchanging line. It uses the same estimate of measurement error as the RCI (and therefore, if you don't trust that estimate, don't use either the RCI or RTI). 

Both the RCI and RTI ask the question: "How likely is this sequence of scores if there is no true change in my data?" The central difference is that the RCI is a two-timepoint instance of the more general RTI. 

## Package goals  

The goal of this package is to provide a simple interface for computing the RCI and RTI given a data set. I intend it for my own personal use and as an educational/demonstration tool in this ongoing research endeavor. 

If you are interested in contributing to this work in any way (including by pointing out mistakes or providing counterarguments), please get in touch with me: [@andrewathan](https://andrewathan.bsky.social) or 
[GitHub](https://github.com/andrewmcaleavey/ReliableTrendIndex).

I think the RTI is a demonstrably better operationalization of reliable change than the RCI in almost any case with more than 2 data points. With two data points, it simplifies to the RCI. The chief added assumption with more time points is that linear change is a reasonable approximation of the true change or that a linear trend is of interest for the data regardless of any curvilinear aspects. So, the RTI assumes you are interested in an overall tendency to increase or decrease over the course of the observation period. 

Nonlinearity of the true change process will make the RTI less sensitive, but still more accurate than the RCI. The RCI does not explicitly pay a penalty for nonlinear changes, it also fails to identify them either. If nonlinearity is a serious concern, the RCI will not be useful due to this reason.   

In general, it is not my opinion that the RTI solves all of the RCI's problems. There are many reasons not to be interested in reliable changes at all. But if you are, the RTI should be a usable solution.  

## Installation  

I am providing this package publicly because it should be available to everyone who is interested in these issues, to discover strengths and weaknesses of this framework. Not because I think it is perfect. That is, errors might be present.  

You can install the development version of ReliableTrendIndex from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("andrewmcaleavey/ReliableTrendIndex")
```

## Usage  

You could use this package just to have a standard way to access RCI and RTI-like functions. They're not complex, and other packages exist to compute them. But it is easier to have them in one standard format than bother with always having to re-write these functions with varying levels of stability on a per-project basis, and putting them together will clarify their similarities.  

However, *this package should be used with caution*. The idea of "reliable" change is questionably justified and has a strong tendency to confuse or mislead. Plus, mistakes may exist, so it is provided as-is.    

Bugs and issues can be reported at [the GitHub repo](https://github.com/andrewmcaleavey/ReliableTrendIndex). If anyone would like to contribute, please be in touch!  

## Theory: Reliable Trend Index (RTI)

### Motivation
In clinical and psychological measurement, we often ask whether an observed *change* in scores is larger than what could be expected from measurement error. The **Reliable Change Index (RCI; Jacobson & Truax, 1991)** addresses this for a **two-timepoint difference score**.  

However, in many contexts we observe a *series* of measurements from a single person (e.g., weekly or daily assessments). In such cases, it is natural to ask whether there is a **reliable linear trend** over time, not just a reliable change between two occasions.  

The **Reliable Trend Index (RTI)** extends the logic of the RCI to this setting: it is a standardized test statistic for the within-person slope in a regression of scores on time, using a reliability-adjusted estimate of the slope’s standard error.

---

### Regression framework
Suppose one person has $n$ observations $y_1,\dots,y_n$ at equally spaced times $t=1,\dots,n$. We fit a simple regression model

$$
y_t = \beta_0 + \beta_1 t + \varepsilon_t,
$$

and test whether the slope $\beta_1$ differs from zero.  

The OLS slope estimate is

$$
\hat\beta_1 = \frac{\sum_{t=1}^n (t-\bar t)(y_t-\bar y)}{\sum_{t=1}^n (t-\bar t)^2}.
$$

Its standard error is

$$
SE(\hat\beta_1) = \frac{\sigma_e}{\sqrt{\sum_{t=1}^n (t-\bar t)^2}},
$$

where $\sigma_e$ is the standard deviation of the residuals.  

---

### Plugging in reliability
If the measurement has reliability $r$ and single-occasion standard deviation $SD$, then the standard deviation of measurement error is

$$
\sigma_e = SD \sqrt{1-r}.
$$

For equally spaced times, the denominator simplifies to

$$
\sum_{t=1}^n (t-\bar t)^2 = \frac{n(n^2-1)}{12}.
$$

Thus,

$$
SE(\hat\beta_1) = SD \sqrt{1-r} \;\sqrt{\frac{12}{n(n^2-1)}}.
$$

---

### Definition of the RTI
The **Reliable Trend Index (RTI)** is defined as the slope estimate divided by this reliability-adjusted standard error:

$$
\text{RTI} = \frac{\hat\beta_1}{SD \sqrt{1-r}\,\sqrt{\tfrac{12}{n(n^2-1)}}}.
$$

This statistic can be interpreted analogously to a $t$-test: values beyond $\pm 1.96$ (approx.) suggest a trend unlikely to be explained by measurement error alone, at the $\alpha=0.05$ level.

---

### Relationship to the Reliable Change Index (RCI)
When $n=2$, the regression slope reduces to the difference between the two observations:

$$
\hat\beta_1 = y_2 - y_1,
$$

and the denominator simplifies to

$$
SE(\hat\beta_1) = SD \sqrt{2(1-r)}.
$$

Therefore,

$$
\text{RTI}_{n=2} = \frac{y_2 - y_1}{SD\sqrt{2(1-r)}},
$$

which is exactly the classical **Reliable Change Index (RCI)**.  

The RCI is thus a special case of the RTI for two timepoints. The RTI generalizes the logic to an arbitrary number of repeated measures, testing whether a *trend* in scores is reliably greater than chance, given the measure’s reliability.  

## Definition  

Jacobson & Truax defined the RCI as $$RC = \frac{x_2 - x_1}{S_{diff}}$$ The term $x_2 - x_1$ is the observed change, which is treated as an unbiased estimate of true change. $S_{diff}$ is the standard error of the difference score, the appropriate error term for a difference score if the assumptions of the model are accurate. Notably, this is based on data external to the person in question. 

The RTI is defined most basically as $$RT = \frac{b}{s_{e}}$$ Here, $b$ is the linear time effect (slope) estimate from a regression of all available data points and $s_{e}$ is the appropriate standard error for that slope, estimated from external data and using the same assumptions as the RCI. However, $s_{e}$ will be smaller than $S_{diff}$ as a function of the number of observations. The logic of this is that the more observations of a series, the less likely a random series is to have a notably increasing or decreasing linear trend so the plausible error decreases.     

There are currently two operationalizations of RTI definitions. The first version of this package used [`metafor`](https://cran.r-project.org/package=metafor) as a backend, which is a flexible approach. The second uses a simpler dependency-free approach to extend the RCI to multiple time points, but lacks some flexibility. Current plan is to use the simpler model for most functions, but this is not implemented yet. The text below only applies to the {`metafor`} approach.

Note that the definition of the RTI is also more obviously model-sensitive than the RCI, because more data points means more options for models (the RCI is model-dependent as well, but its model makes intuitive sense given a difference score). Depending on the number of observations, the RTI will be defined differently:  

  - **For 2 observations**, the RTI is the same as the RCI. To compute a linear regression, the RTI as implemented here treats both observations and the difference score as fixed values. This means that the slope of the best fitting line must cross through both observations. Put differently, the best available estimate of $b$ is simply $x_2 - x_1$. The standard error of a difference score ($S_{diff}$) is the appropriate standard error for this case. Thus, the entire model simplifies to the RCI.   
  - **For 3 observations**, there are two options. We could treat all three observations as fixed, but this means that there is no gain from the RTI compared to the RCI. The model will simply take the first and last observation, which is equivalent to the RCI. Instead, by default, the RTI treats the intercept as fixed but estimates a slope using the other observations. This allows for uncertainty in the slope estimation, though it implies that slope comes from an unrealistically-accurately measured first observation. The slope will not differ if a different point were chosen as the fixed point.     
  - **For 4 or more observations**, the RTI estimates both a slope and intercept. This should be the most generalizable method for trend identification. It does not prioritize any given observation(s), treating them all with equal potential error. This does mean that the slope will likely not pass directly through any specific observations: the most likely true state will be estimated to not coincide with the observations, because those observations have error.  

## Example

```{r load package}
library(ReliableTrendIndex)
```

### One person RCI  

To calculate the RCI for an individual, we can do the following:  

```{r example with one person}
## here is the example case provided by Jacobson & Truax (1991): 
knitr::kable(jt_example_data_1)
```

We will use that data to compute a simple RCI.  

The difference score for this person is `47.5 - 32.5`, or `15`. 

The scale's standard error of the difference score, as computed by the authors, is `4.74`. 

```{r}
jt_rci_calc(difference = 15, sdiff = 4.74)
```

That value, 3.16, is the person's RCI, according to Jacobson & Truax. Since it is greater than their suggested cutpoint of 1.96 (which conveys 5% chance of Type I error under the assumptions of the RCI), we would conclude that this change is "reliable."

### One person RTI  

If we had assessed a person more than twice, it would be nice to incorporate all of the information we have about them, rather than just the first and last (or any two observations in isolation). For the purpose of an example, we will use the fabricated height data from McAleavey (2022).  

```{r}
knitr::kable(mac_height)
```

That data shows a difference score of `1` cm. From the example text, we know that the standard error of the difference is `.7071068` and the standard error of measurement is `0.5`. 

We could simply compute the RCI for this individual by using their first and last observations, which is commonly done in routine care and clinical trials analysis of clinically significant change:  

```{r}
jt_rci_calc(difference = 1, sdiff = .7071068)
```

That value, 1.414, is not greater than 1.96, so we would conclude that the change from pre-post is not reliable. The `ReliableTrendIndex` term for this is `Less than reliable` as opposed to `No Change` in order to indicate that the method could not specifically identify the change score. 

However, the RTI would incorporate all six measurements. It is a waste to ignore two-thirds of our information here.  

In its most simple form, the RTI takes the RCI's observed difference score question (How likely is this difference score, if the true change was actually 0?) and extends it to the overall sequence of scores (How likely is this sequence of scores if the true linear trend was actually 0?). 

The simplest function to use for a single person - whether computing the RCI and/or RTI - is `rti()`. 

```{r}
mac_rti <- rti(mac_height$obs, sdiff = .707)
```


Note that it gave us this message on screen: `"More than two values provided, assuming they are evenly spaced in time."`. The function `rti()` is meant to be *simple*, so it is not meant to be used with uneven assessment spacing. Other methods allow more flexibility at the cost of increase complexity. 

The object `mac_rti` is of the class `reliableTrend`. It contains a lot of information and can be viewed with `print.reliableTrend()` (which is also called by just the object name):

```{r}
print(mac_rti)
```
 
An additional text summary can be found using `summary()`. 

```{r}
summary(mac_rti)
```

So now we can clearly see that the RCI, by ignoring the interim measurements and using a strong cutoff, would not detect a reliable change but the RTI would. Note that the summary also tells us the probability of the result being in its most likely direction. It is not uncommon for large probabilities like 80-90% to be considered "Unreliable" due to the conservative Type I error approach of the RCI.

You might want to visualize this to see what it's doing. Try `plot.reliableTrend()` (or just `plot()`). 

```{r}
plot(mac_rti) 
```

Notice that the trend line is going up, and the 95% CI for the RTI (shaded region) is more precise than the 95% CI for individual observations (error bars), which is what the RCI on its own relies on.    

The output of `plot.reliableTrend()` is a `ggplot2` object, so you can modify it in that ecosystem: 

```{r}
plot(mac_rti) +
  ggplot2::labs(title = "Mac's height example",
                y = "Height", 
                x = "Occasion") +
  ggplot2::theme_dark()
```


## Analysis of complete data sets  

More complete functions to analyze a complete data set are also provided in the function documentation and an initial `vignette("Introduction_to_ReliableTrendIndex")`. Complete documentation to come.

## License

This package is licensed under the GNU Affero General Public License (AGPL) 3.  
