---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ReliableTrendIndex

<!-- badges: start -->
<!-- badges: end -->

An R package for analysis of "reliable" changes.

PLEASE DO NOT USE RELIABLE CHANGE OR THIS PACKAGE FOR ANYTHING WITH MEANINGFUL CONSEQUENCES. 

The idea of reliable change is extremely questionable, and there are almost certainly better ways to investigate the questions in which you are interested. 

The goal of this package is to provide a simple interface for computing the RCI and RTI given a data set. I intend it for my own personal use and as an educational/demonstration tool in this ongoing research endeavor. 

## Installation

I don't recommend installing this package. I am providing it publicly because it should be available to everyone who is interested in these issues, to discover strengths and weaknesses of this framework. 

You can install the development version of ReliableTrendIndex from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("andrewmcaleavey/ReliableTrendIndex")
```

## Usage

You could use this package just to have a standard way to access RCI and RTI-like functions. They're not complex, other packages exist that compute them, but it is easier to have them in one standard format than bother with always having to re-write these functions with varying levels of stability on a per-project basis.  

I repeat, however, that *I don't really think these functions should be used for any clinically-meaningful application* except with extreme caution. The idea of "reliable" change is questionably justified and has a strong tendency to confuse or mislead users.  

Bugs and issues can be reported at [the GitHub repo](https://github.com/andrewmcaleavey/ReliableTrendIndex). If anyone would like to contribute, please be in touch!  

## Example

```{r load package}
library(ReliableTrendIndex)
```

### One person RCI  

To calculate the RCI for an individual, we can do the following:  

```{r example with one person}
## here is the example case provided by Jacobson & Truax (1991): 
knitr::kable(jt_example_data_1)
```

We will use that data to compute a simple RCI.  

The difference score for this person is `47.5 - 32.5`, or `15`. 

The scale's standard error of the difference score, as computed by the authors, is `4.74`. 

```{r}
jt_rci_calc(difference = 15, sdiff = 4.74)
```

That value, 3.16, is the person's RCI, according to Jacobson & Truax. Since it is greater than their suggested cutpoint of 1.96 (which conveys 5% chance of Type I error under the assumptions of the RCI), we would conclude that this change is "reliable."

### One person RTI  

If we had assessed a person more than twice, it would be nice to incorporate all of the information we have about them, rather than just the first and last (or any two observations in isolation). For the purpose of an example, we will use the fabricated height data from McAleavey (2022).  

```{r}
knitr::kable(mac_height)
```

That data shows a difference score of `1` cm. From the example text, we know that the standard error of the difference is `.7071068` and the standard error of measurement is `0.5`. 

We could simply compute the RCI for this individual by using their first and last observations, which is commonly done in routine care and clinical trials analysis of clinically significant change:  

```{r}
jt_rci_calc(difference = 1, sdiff = .7071068)
```

That value, 1.414, is not greater than 1.96, so we would conclude that the change from pre-post is not reliable. The `ReliableTrendIndex` term for this is `Unspecified` as opposed to `No Change` in order to indicate that the method could not specifically identify the change score. 

However, the RTI would incorporate all six measurements. It is a waste to ignore two-thirds of our information here.  

In its most simple form, the RTI takes the RCI's observed difference score question (How likely is this difference score, if the true change was actually 0?) and extends it to the overall sequence of scores (How likely is this sequence of scores if the true linear trend was actually 0?). 

The simplest function to use for a single person - whether computing the RCI and/or RTI - is `rti()`. 

```{r}
mac_rti <- rti(mac_height$obs, sdiff = .707)
```


Note that it gave us this message on screen: `"More than two values provided, assuming they are evenly spaced in time."`. The function `rti`()` is meant to be *simple*, so it is not meant to be used with uneven assessment spacing. Other methods allow more flexibility at the cost of increase complexity. 

The object `mac_rti` is of the class `reliableTrend`. It contains a lot of information and can be viewed with `print.reliableTrend()` (which is also called by just the object name):

```{r}
print(mac_rti)
```
 
An additional text summary can be found using `summary()`. 

```{r}
summary(mac_rti)
```

So now we can clearly see that the RCI, by ignoring the interim measurements and using a strong cutoff, would not detect a reliable change but the RTI would. 

You might want to visualize this to see what it's doing. Try `forest_to_reg_plot()`. 

```{r}
forest_to_reg_plot(mac_rti$rmaObj, 
                   StError = sqrt(mac_rti$error_var)) +
  ggplot2::labs(title = "Mac's height example",
         y = "Height")
```

Notice that the trend line is going up, and the 95% CI for the RTI (shaded region) is more precise than the 95% CI for individual observations (error bars), which is what the RCI on its own does.    

## Analysis of complete data sets  

More complete functions to analyze a complete data set are also provided in the function documentation and an initial `vignette("Introduction_to_ReliableTrendIndex")`. Complete documentation to come!

## License

This package is licensed under the GNU Affero General Public License (AGPL) 3.  
