---
title: "Introduction_to_ReliableTrendIndex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction_to_ReliableTrendIndex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ReliableTrendIndex)
```

I hope no one ever reads this. Except me, probably, when I want to do revisions to some paper later on. 

The package `{ReliableTrendIndex}` is a way to contain a large group of disorganized analysis functions related to the malformed concept of reliable change. There are a few key functions. 

### `reliableTrend()` and the `reliableTrend` class  

The function `reliableTrend()` creates objects of class `reliableTrend`. This class is just a list with fixed, named entries. You can read about it in `?reliableTrend()`.  

### Meta-analysis backbone  

The functions are mostly wrappers around functions from the `{metafor}` package, available on CRAN. The chief of these is `metafor::rma()`, which conducts simple univariate meta-analyses. Read about it at `?metafor::rma()`. It was worth developing wrapper functions for me because **A)** those functions are much more complex than the RCI/RTI I am working with here, and **B)** the variables in that package are named appropriately nonspecific things for meta-analysis, but overly complex things for me. 

## Workflows  

Some basic usage examples are provided in `?reliableTrend`, `?simple_rma`, `?rti_to_df` and others. 
The two essential use cases are when you have one person, and when you have more than one person. 

### One person workflows  

Obviously, this is the simple case. 

### Multi-person workflows  

In the case of analyzing a lot of data simultaneously (e.g., in a research project or simulation study), the simplest way to think about it is doing the previous case iteratively. However, you might have one, two, or five hundred observations for a given person - it would be ideal to have an analysis system that could handle all of those cases. 

```{r example of multiple ppl}
test4 <- simulated_data %>%
  split(.$id) %>%
  purrr::map(~ simple_rma(., error_var = .2^2)) %>%
  purrr::map(reliableTrend) %>%
  purrr::map_dfr(rti_to_df) %>% 
  mutate(id = simulated_data %>% group_by(id) %>% slice(1) %>% pull(id)) %>% 
  right_join(simulated_data)

table(test4$category.RCI, test4$true_change)
table(test4$category.RTI, test4$true_change)
```

