---
title: "Migrating to the supported ReliableTrendIndex API"
output:
  rmarkdown::html_vignette:
    math_method: mathjax
    mathjax: false
    pandoc_args: ["--math-method=mathjax"]
vignette: >
  %\VignetteIndexEntry{Migrating to the supported ReliableTrendIndex API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Supported API

New analyses should use three functions:

- `rti()` for one series of repeated scores;
- `rci()` for a two-occasion change score; and
- `rti_by()` for long-format data with one result per person or group.

Each RTI workflow requires externally justified measurement-error information.
The preferred RTI inputs are `sd` and `r`; `rti()` also accepts externally
derived `sem` or `sdiff`.

## Replacing historical helpers

| Historical call | Replacement |
| --- | --- |
| `reliableTrend(values, time, sem)` | `rti(y = values, t = time, sem = sem)` |
| `slope_se_reliability(y, r, time, sd_single)` | `rti(y, t = time, sd = sd_single, r = r)` |
| `rti_by_person(...)` | `rti_by(data, id, time, y, sd, r)` |
| `jt_rci_calc(...)` or `rci_from_scores(...)` | `rci(...)` |
| `rti_calc_simple(values, sem)` | `rti(y = values, sem = sem)` |

The historical helpers remain functional but issue a deprecation warning. They
are planned to remain available until at least version 1.0.0.

## Example

```{r}
library(ReliableTrendIndex)

scores <- c(10, 12, 11, 15)
times <- c(0, 1, 3, 4)

fit <- rti(y = scores, t = times, sd = 8, r = 0.85)
fit

rci(difference = scores[length(scores)] - scores[1],
    sdiff = 8 * sqrt(2 * (1 - 0.85)))
```

## Preserved legacy output

Compatibility wrappers retain their historical return classes and fields. For
example, `slope_se_reliability()` still returns a `slopeSErel` object and
`rti_calc_simple()` still returns its `rmaObj`, `error_var`, and `fit` fields.
Use the supported API for new scripts so that code remains valid after the
planned 1.0.0 lifecycle transition.
